---
title: "Part 1 Scripts"
author: "M.Soria & A.Dehlia"
date: "5/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to use this file

This file is a companion to the **part1.R** located at the Scripts folder of this repository. This file will give a brief description for each script together with its input and output.


### Read a tab separated values (.tsv) file 
A .tsv can be read using the function read.csv() by setting the attribute, *sep*, to the tab value seen in the code below. The *row.names* attribute takes the value of the column, GeneID, and turns into the row names of the data frome. The head() function set to n=6 shows the first 6 items in the file.

**Function:** read.csv() <br />
**Input:** .tsv file <br />
**Output:** data.frame

```{r, echo=TRUE}
gene_expression_data <- read.csv("Data/part1_gene_expression.tsv", 
                                 sep = '\t', row.names = "GeneID")
head(gene_expression_data, n=6)
```


### Add a new column and calculate the mean of two columns 
The following script assigns a new column name *Mean* to the data frame *gene_expression_data*. The dollar sign ($) is used to denote that Mean is a colum of gene_expression_data and the symbol *<-* is an R syntax used to assign values to a variable. <br />

The rowMeans() function takes the values of a set of columns from a data frame and calculates the mean for each row. In this script, the columns are denoted by *[,1:2]* which includes the first two columns of the data frame.

**Function:** rowMeans()<br />
**Input:** values for the new column <br />
**Output:** data.frame (+ new column)

```{r, echo=TRUE}
gene_expression_data$Mean <- rowMeans(gene_expression_data[,1:2])
head(gene_expression_data, 6)
```


### List the top 10 highest mean values  
This script first orders the values in the Mean column of the data frame gene_expression_data from greatest to least (denoted by the "-" sign), and stores the ordered values in the gene_ordered variable. The head() function with the attribute, 10, displays the top ten rows with the highest mean values.


**Function:** order()<br />
**Input:** values from column <br />
**Output:** data.frame

```{r, echo=TRUE}
gene_ordered <- gene_expression_data[order(-gene_expression_data$Mean),]
head(gene_ordered, 10)
```


### Determine the number of genes with a mean less than 10 
This script uses a conditional expression to filter out the mean values less than 10. 


**Input:** values from column, conditional statement <br />
**Output:** data.frame


The nrow() function counts the number of rows that satisfies the condition. 


**Function:** nrow()<br />
**Input:** data.frame <br />
**Output:** int

```{r, echo=TRUE}
mean_lessthan10 <- gene_expression_data[(gene_expression_data[,3]<10),]
nrow(mean_lessthan10)
```


### Make a histogram plot of the mean values saved in png format
The hist() function creates a histogram for the Mean column. The argument *xlab* creates the label for x-axis and *main* creats the title of the graph. 


**Function:** hist()<br />
**Input:** vector of values <br />
**Output:** histogram


```{r, echo=TRUE}
hist(gene_expression_data$Mean, xlab = "Mean", main = "Histogram of Gene Expression Data Mean")
```

The png() creates a .png file and the argument *filename* can be used to input the filename and the location where the image is to be saved. The function dev.off() shuts down the png function, telling R that the file is finished and is ready to be saved. The plot function shoud be run in between png() and dev.off().


The function ftable() creates a frequency table for values in a column.

The function mean() calculates the mean value of a list of values.


### Read a comma separated values (.csv) files
A .csv can be read using the function read.csv().


**Function:** read.csv()<br />
**Input:** .csv file <br />
**Output:** data.frame

The function colnames() gets or sets the name of the column of this data.frame.

**Function:** colnames()<br />
**Input:** data.frame <br />
**Output:** list of characters

```{r, echo=TRUE}
growth_data <- read.csv("~/projects/SLE712-Assignment-3/Data/part1_growth_data.csv",
                        header = TRUE, stringsAsFactors = FALSE)

column <- colnames(growth_data)
column
```


### Calculate the standard deviation of a vector of values

The function sd() is used to calculate the standard deviation of a list of values. 


**Function:** sd()<br />
**Input:** numeric vector <br />
**Output:** number

```{r, echo=TRUE}
ne <- subset(growth_data, Site=="northeast")
sd(ne$Circumf_2004_cm)
```


### Make a boxplot

The function boxplot() is used to create a  box-and-whisker plot from a set of values. 


**Function:** boxplot()<br />
**Input:** data.frame or list <br />
**Output:** boxplot

```{r, echo=TRUE}
sw <- subset(growth_data, Site == "southwest")
boxplot(ne$Circumf_2004_cm, ne$Circumf_2019_cm, sw$Circumf_2004_cm, sw$Circumf_2019_cm,
        names = c("NE 2004", "NE 2019", "SW 2004","SW 2019"),
        ylab= "Circumfrence (cm)" , xlab = "Sites and years" ,
        main = "Growth at two different sites during 2004 and 2019", col= "green")
```


### t.test and wilcox.test functions

The t.test() or Student's t-Test function is used to make sample t-tests on a group of data. 

**Function:** t.test()<br />
**Input:** numeric vector <br />
**Output:** List

```{r, echo=TRUE}
# Mean growth for Northeast data
ne$Circumf_2019_cm - ne$Circumf_2009_cm
ne$growth <- (ne$Circumf_2019_cm - ne$Circumf_2009_cm)

# Mean growth for Southwest data
sw$Circumf_2019_cm - sw$Circumf_2009_cm
sw$growth <- (sw$Circumf_2019_cm - sw$Circumf_2009_cm)

t_test <- t.test(ne$growth,sw$growth)
t_test
```

The wilcox.test() or Wilcoxon Rank Sum and Signed Rank Tests function is used to make sample Wilcoxon on a group of data. 


**Function:** wilcox.test()<br />
**Input:** numeric vector <br />
**Output:** List

```{r, echo=TRUE}
# Mean growth for Northeast data
ne$Circumf_2019_cm - ne$Circumf_2009_cm
ne$growth <- (ne$Circumf_2019_cm - ne$Circumf_2009_cm)

# Mean growth for Southwest data
sw$Circumf_2019_cm - sw$Circumf_2009_cm
sw$growth <- (sw$Circumf_2019_cm - sw$Circumf_2009_cm)

wilcox_test <- wilcox.test(ne$growth, sw$growth)
wilcox_test
```


## Additional Methods to determine details of a function
R has a built in library of decriptions for its functions. To access these descriptions, use the *?* symbol followed by the name of the function and run that within the R script or the console. The descriptions will be shown in the Help tab located by default at the lower right corner of the R environment.

```{r, echo=TRUE}
?mean 
```